---
title: "p8105_mtp_amv2187"
author: "Alyssa Vanderbeek"
date: "26 October 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(patchwork)

theme_set(theme_bw())
```

```{r}
accel_data = read.csv('./data/accel_data.csv') %>% # import
  janitor::clean_names() %>% # clean names
  gather(key = minute, value = activity_level, contains('activity')) %>% # tidy activity columns into single variable
  mutate(minute = as.integer(str_sub(minute, 10)), # adjust activity period to numeric only
         day = fct_relevel(day, 'Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday'), # relevel day of the week to go from Sunday to Saturday
         hour = ceiling(minute / 60)) %>% # the device records activity units for every minute of the day, so we can break this up into hours of the day, starting at midnight
  arrange(week, day) # arrange df by week and day

str(accel_data)

# aggregate daily total activity counts
agg_days = accel_data %>%
  group_by(week, day) %>% # group by week and day
  summarise(total_activity = sum(activity_level)) %>% # get total activity for each day
  ungroup %>%
  mutate(order = row_number()) # order variable to plot later

str(agg_days)

```

## Overview

Here I examine activity levels of a 63-year-old man over the course of several months (47 weeks). The individual has a BMI of 25, and was admitted to the CUMC Advanced Cardiac Care Center for congestive heart failure before participating in this study. The goal of the research is to understand whether activity level changes over time after a major cardiac event, and whether there is potential to gather information about health events from the data surrounding them.

The dataset is comprised of information collected every minute over 47 weeks (329 days, `r nrow(accel_data)` minutes).  

Activity level is measured in voltage, where higher counts indicate higher levels of activity. Of note, there was no missing data per se, but we can understand an activity count of 1 to mean that no activity was recorded in that minute; either the participant was not active or not wearing the device. For example, an entire day of recorded 1 values might be understood as a day on which the man did not wear the device. There were `r agg_days %>% filter(total_activity == 1440) %>% count` such days recorded. 



## Analysis

Upon receiving and tidying the data, I looked at the distribution of activity level and the relationship between activity level and time. There are seemingly many outliers - in particular, disproportionately high minute-to-minute activity counts. Taking the log of the minute-to-minute counts reduces this skewness, but aggregating by hour or day does as well.


### Daily activity over time

We are interested to know whether the man's activity levels increased over time. I fit a linear model with the day (1 to 329) as the predictor, which suggests that activity level did on average increase over the course of the observed months (p<0.0001; Figure 1). This trend makes sense in context; activity was restricted after the cardiac event (as is commonly recommended), and was then increased as time went on. It would be interesting to know the man's activity levels prior to the event as an informative contrast. 

```{r}
# linear model to test whether activity changed over time. Significant positive slope says activity increased; significant negative intercept says acitvity decreased. Lack of significance fails to reject the null hypothesis that day over the course of the months of wear does not predict activity level.
summary(lm(agg_days$total_activity ~ agg_days$order))

# scatter plot of total activity by day
agg_days %>%
  ggplot(aes(x = order, y = total_activity)) +
  geom_point() + 
  geom_smooth(method = 'lm', se = F)  + # fit linear model, hide SE regions
  labs(
    title = 'Daily activity level over time',
    y = 'Total activity level',
    x = 'Time (days)',
    caption = 'Figure 1'
  )
```

Since we also have information on the day of the week (DOW), we can test the hypothesis that DOW also predicts activity level. An ANOVA test indicates that there is little evidence to suggest that activity level varies by DOW (Figure 2). This holds true when accounting for time (Figure 3). Interestingly, Saturday goes from being, on average, the least active DOW to the most active by the end of the study period.

```{r}
# ANOVA testing relationship between activity level and day of the week
anova(lm(agg_days$total_activity ~ agg_days$day)) # ANOVA test of difference between means

# boxplot of total daily activity by day of the week
boxplot_dow = agg_days %>%
  ggplot(aes(x = day, y = total_activity, fill = day)) +
  geom_boxplot() +
  labs(
    title = 'Daily activity level by day of the week',
    x = 'Day of the week',
    y = "Total activity level",
    caption = 'Figure 2'
  ) + 
  theme(legend.position = 'none',
        title = element_text(size = 8),
        axis.text.x = element_text(angle = 60, hjust = 1)) + # angle DOW on x-axis for readability
  viridis::scale_fill_viridis(discrete = T)

# ANOVA testing relationship between activity level and day of the week over time
anova(lm(agg_days$total_activity ~ agg_days$day + agg_days$order)) # ANOVA test of difference between means

# linear fit for total daily activity by day of the week
scatter_dow_time = agg_days %>%
  ggplot(aes(x = order, y = total_activity, color = day)) +
  geom_point() +
  geom_smooth(method = 'lm', se = F) +
  labs(
    title = "Daily activity level by day of the week over time",
    x = 'Time (days)',
    y = 'Total activity level',
    caption = 'Figure 3'
  ) +
  theme(legend.position = 'none',
        title = element_text(size = 8)) + 
  viridis::scale_color_viridis(
    name = 'Day of the week',
    discrete = T
  )

boxplot_dow + scatter_dow_time

```

### Activity over the course of the day

I examined the distribution of activity by week, day of the week, and hour of day, and found that the participant was most active during his waking hours, between 7 am and 11 pm, approximately (Figure 4). This trend does not vary much by DOW (Figure 4), but we can see a general increase in activity during waking hours over time (Figure 5, in agreement with prior linear model). 

```{r, warning=F}
# daily 24-hour activity profile (spaghetti plot)
#hourly = 
accel_data %>%
  group_by(week, day, hour) %>% # group by week, day, and hour
  summarise(total_activity_hr = sum(activity_level)) %>% # get total activity for each hour of every day
  ungroup %>%
  mutate(day_over_time = ceiling(row_number()/24)) %>% # variable to group by day 1:329
  ggplot(aes(y = total_activity_hr, x = hour)) +
  geom_line(aes(group = day_over_time), alpha = 0.3) + # plot the 24-hr activity for each day
  geom_smooth(aes(group = day, color = day), se = F, alpha = 0.3, size = 0.5) + # add smoothed lines for day of the week
  labs(
    title = '24-hour activity for all recorded days',
    x = 'Hour of day (beginning at midnight)',
    y = 'Activity Level',
    color = 'Aggregated trends \nby day of the week',
    caption = 'Figure 4'
  ) 


# 24-hour activity profile over time for each week
accel_data %>%
  group_by(week, hour) %>% # group by week and hour
  summarise(total_activity_hr = sum(activity_level)) %>% # get total activity for each hour of every day
  ungroup %>%
  ggplot(aes(x = week, y = hour)) +
  geom_point(aes(size = total_activity_hr), alpha = 0.5) +
  labs(
    title = 'Hourly activity level over time',
    y = 'Hour of day (beginning at midnight)',
    x = 'Time (weeks)',
    caption = 'Figure 5. The size of the points indicates the amount of activity for a certain hour of the day (y-axis) \n in a given week (x-axis);smaller points indicate less activity and larger points indicate more.',
    size = 'Activity level'
  ) + 
  theme(legend.position = 'none') +
  viridis::scale_color_viridis(discrete = T)

```


```{r}
wordcountaddin::text_stats("p8105_mtp_amv2187.Rmd")
```


